Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-15 paper.tex
Updates to early recomb from Ana

\section{Introduction}
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}),
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.

Recombination was challenging to detect early in the SARS-CoV-2 pandemic due to
gradual accumulation of genetic variation. In late 2020, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations, increased the
detectability of recombination. By searching for genomes containing segments
from different VoC designations, multiple recombinant lineages have been
detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
and Epsilon \citep{Wertheim2022-hj}, and Alpha and
Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for recombination
\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of
onward transmission (more than five descendants) and monophyly ($>70\%$
bootstrap support) are eligible for a Pango lineage
designation~\citep{Rambaut2020-dw}, starting with the letter X
% FIXME: should this be a citation?
(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant
Lineages'' 2021).

@@

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
development of the ``succinct tree sequence'' data
structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
encoding of large ARGs, and this representation should be well-suited to storing
recombinant viral genealogies, even those with millions of sampled sequence,
such as SARS-CoV-2 (with currently over 15 million sequences in the GISAID database
\citep{Shu2017-hp}). Moreover the \emph{tskit} software library, which implements this
data structure, enables efficient manipulation of the stored genealogies and is
integrated with a suite of open-source software tools to simulate, analyse, and
infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

@@

Here, we present a preliminary version of a novel scalable methodology
(\emph{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
and additionally takes advantage of known collection dates to infer ARGs from millions
of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
we demonstrate that sc2ts can be used to simultaneously (1) detect putative SARS-CoV-2
recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
(3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
\emph{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants
and to jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at scale.

@@

The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
% TODO more discussion here, making sure we explain anything that might not be
% obvious.
Most of the nodes in the ARGs represent sample genomes
(Wide ARG: 87\%, Long ARG: 84\%), but the inferred ancestral sequences
represented by the nodes inferred from daily sample clusters
by UPGMA (Section~\ref{sec:sample-cluster-tree-inference})
and parsimony heuristics (Section~\ref{sec:parsimony-heuristics})
contribute significantly.
Table~\ref{tab:args} shows the counts of edges (see
Section~\ref{sec:args} for a technical definition) in the ARGs
categorised by parent node type, and demonstrates that each
of these contributes significantly to the overall topology.
In particular, we can see that approximately 40\% of the edges
in both ARGs involve inheriting directly from sampled genomes.
Both ARGs contain 29,422 sites (\ref{sec:data_preprocessing}),
and a large number of mutations.
The average number of mutations per-site is high, although much
of this may be explained by outlier sites with artefactually high
mutation counts (see Figure~\ref{fig:breakpoint-distribution}).
Despite this, however, the number of mutations per node
is small. In the Wide ARG, for example, we have a mean of 0.77
mutations per sample, demonstrating that new samples are generally
well represented by the it. See Section~\ref{sec:mutation_spectrum}
for more analysis of the patterns of inferred mutations.
Recombination plays a relatively minor role, with $<0.3\%$
of nodes in the ARGs representing inferred recombination events.
We analyse these events in Sections~\ref{sec:jackson_recombs},
\ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},
and \ref{sec:pango_x_lineages}.

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing a local tree from the Wide ARG
(sampled to mid-2021) and an ``all-time'' global NextStrain tree
(downloaded on January 21, 2023).
Phylogenies are pruned down to those samples present as tips in both datasets. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated using the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

A classic tree-based summary of SARS-CoV2 ancestry is provided
by the NextStrain project \citep{Hadfield2018-ef}. The trees
available from Nextstrain are based on small subsamples of the
dataset, and early in the pandemic tend to be restricted to
a sample of strains which are not thought to be recombinants.
For validation purposes, we our compare our ARGs with
a downloaded NextStrain tree, restricted to the time period
covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
the \texttt{sc2ts} ARGs to a ``backbone'' containing only samples present
the NextStrain tree. This results in a small set of shared
samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
assigned to Pango recombinant lineages by Nextclade (see Section
\ref{sec:pango_x_lineages}).

Our \texttt{sc2ts} backbone phylogenies do contain small amounts of recombination:
there are 7 recombination nodes in the backbone of the Wide ARG and 8 in the backbone
of the Long ARG. However, the backbones remain extremely tree-like, as these
recombination events involve minor, local topological rearrangements. In particular,
all involve recombination between closely related lineages (e.g.\ sibling lineages)
and a number even represent recombination within a single lineage, in which the parent
lineages above a recombination event coalesce with each other before coalescing
with any other lineages in the phylogeny. These could either represent true
recombination events between closely related strains, which have previously gone unnoticed,
or could be artifacts of the current \texttt{sc2ts} method and its parameterization.

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from NextStrain
% We give the date below in the caption - we could probably remove if either from there
% or from here (I suggest from here)
(downloaded on January 21, 2023).
We illustrate the backbone phylogeny by visualizing
a single tree in the middle of the viral genome
since other regions of the
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

\subsection{Early recombinants}
\label{sec:jackson_recombs}
% FIXME: this is rubbish, can someone do better?
Although coronaviruses are generally known to recombine [CITATION?]
it was not clear that SARS-CoV-2 recombinants with ongoing transmission
existed early in the pandemic.
\citet{Jackson2021-ik} performed a careful analysis of sequences
circulating in the UK in late 2020 to early 2021, and found
evidence of multiple independent recombination events and onward
transmission.
By searching for samples combining genomic segments from Alpha (B.1.1.7) and
from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
they found 16 recombinant sequences from 8 putative
origins (groups A to D and four singletons).
These findings are closely replicated in both the Wide and Long ARGs.

@@

Both the wide and long ARGs contain a large number of recombination nodes (4123
and 2078 respectively; Supplementary Tables 1 and 2). Each of these nodes were
inferred as a result of matching a specific recombinant sample into the ARG.
However, ~63\% of recombination nodes are ancestral to only a single
recombinant sample; we exclude these "singleton recombinants" from subsequent
analyses as they may have arisen from sequencing of mixed samples (e.g., from
co-infection or contamination), and in any case, never result in onward
transmission.  This leaves 1522 and 763 recombination nodes in the wide and
long ARGs, respectively. In their analysis using RIPPLES,
\cite{Turakhia2022-it} detected 589 non-singleton recombinants.

@@

The date of common ancestry for the two parents of recombinant lineages is
concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
largely due to a small handful of MRCAs which are shared between a large number
of recombination events; the top four are indicated by dotted lines in the
figure. These shared MRCAs are near the root of large expansions of certain
lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first Omicron wave;
and BA.2: second Omicron wave), as evidenced by the number of samples with that
Pango designation that trace back to these ancestral nodes (listed as
percentages on the plot). The specific MRCA nodes to which many recombinants
trace frequently represent very large polytomies in the tree (at the extreme,
2,872 immediate children). This is likely to indicate a rapid and under-sampled
expansion of a clade in the SARS-CoV-2 genealogy.

@@

\subsubsection{Wide ARG}
%% Quick overview of the wide ARG.
In the wide ARG, 44
samples are designated as XA by both Nextclade and GISAID, 237 are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
designated as XB by any method have one or more recombination nodes in their
ancestry.

@@

The remaining remaining 630 Pango recombinant samples trace back to 50
different recombination nodes. We examine the inferred inheritance structure in
the following subsections.

@@

A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
This shows a subgraph of the long ARG, in which paths have been traced from all
Nextclade-identified XA samples to the closest other sample strains in our dataset.
Dotted lines show where this subgraph links to other nodes within the ARG, and above
recombination nodes only ancestral nodes are shown (i.e. we do not descent to include
cousins of recombinant parents).
It is clear from the subgraph that all the XA samples trace to a single originating
recombination node (plotted in black), with a breakpoint position around 22227bp.

@@

A more complex single-origin case is Figure~\ref{fig:pango-simple-origin-graph}B shows a case of another
Pango lineage, XAG, whose Nextclade-designated samples also trace to a
single originating recombination. This is a more intricate example, as the same origin
is also shared with all samples designated XAA, and some, but not all, of the
samples identified as XAB, XQ, and XU by Nextclade. Note, however, that if the GISAID
designations are used, many of the samples marked here as XAB are reclassified
as BA.2, and XAG becomes fully monophyletic (supplementary figure
\ref{fig:pango_XAG_gisaid_graph}, which also gives identifying labels to the
sample nodes, convertable to EPI\_ISL identifiers using
Supplementary table XXX )

Supplementary Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
Pango recombinant lineages that trace back to a single recombination node in
the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
monophyletic clades, in which all of the samples given that Pango designation
descend from a single recombination node.

[Discuss remaining]

There is generally good agreement on the imputed parental lineages from the ARG
and the official Pango designations. [Discuss more]

\subsubsection{Multiple origins}
The ARG inference has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. In this section we examine the Pango X lineage
samples that are inferred to have multiple recombinant origins (i.e., trace
back to independent recombination nodes in the ARG).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL 8514045 (sampled in Turkey and dated 2021-12-30).
These two samples differ by 35 mutations; considering this, plus the geographical
locations and time between the samples,
it does seem possible that they represent true independent origins of XD.
Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs
with sample identifiers and coloured by GISAID Pango labels (however, no
samples are designated by GISAID as XD in the long ARG).

Another X lineage with multiple originating events is XE. In this case, however,
there are a large number of independent events. These

\ref{tab:multiple_origins_table} summarises other Pango recombinants which
we infer to have multiple origins. [TODO: needs more detail]
As with XD, none of the samples in the long ARG are identified as XE by GISAID.

\subsubsection{Complex origins}
As well as cases where single recombination events explain an X lineage,
such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
Pango X lineages involve more complex origins,
in which multiple non-independent (``nested'') recombination events are involved.
In these cases, the descendants of the original recombination
frequently include samples which have not been previously identified as recombinants.
An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are mostly not
identified as recombinants by Nextclade, and plotted in blue. A similar
pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
Note that the descendants of nested recombination events include only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Investigating examples of complex recombinant origins, and identifying which
(if any) of the nested recombination events may be artefactual, is an important
area of future research.

>>>>

\section{Introduction}
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.

The lack of genetic variation made recombination challenging to detect early in the SARS-CoV-2 pandemic. In late 2020, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations, increased the
detectability of recombination. By searching for genomes containing segments
from different VoC designations, multiple recombinant lineages have been
detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
and Epsilon \citep{Wertheim2022-hj}, and Alpha and
Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for recombination
\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of
onward transmission (more than five descendants) and monophyly ($>70\%$
bootstrap support) are eligible for a Pango lineage
designation~\citep{Rambaut2020-dw}, starting with the letter X
% FIXME: should this be a citation?
(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant
Lineages'' 2021).

@@

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
development of the ``succinct tree sequence'' data
structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
encoding of large ARGs. This representation should be well-suited to storing
large recombinant viral genealogies such as SARS-CoV-2,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}). Moreover the \emph{tskit} software library, which implements this
data structure, enables efficient manipulation of the stored genealogies and is
integrated with a suite of open-source software tools to simulate, analyse, and
infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

@@

Here, we present a preliminary version of a novel scalable methodology
(\emph{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
and additionally takes advantage of known collection dates to infer ARGs from millions
of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
we demonstrate that \emph{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
(3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
\emph{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
\emph{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
and to jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at scale.

@@

The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
% TODO more discussion here, making sure we explain anything that might not be
% obvious.
Most of the nodes in the ARGs represent sample genomes
(Wide ARG: 87\%, Long ARG: 84\%), but the inferred ancestral sequences
represented by the nodes inferred from daily sample clusters
by UPGMA (Section~\ref{sec:sample-cluster-tree-inference})
and parsimony heuristics (Section~\ref{sec:parsimony-heuristics})
contribute significantly.
Table~\ref{tab:args} shows the counts of edges (see
Section~\ref{sec:args} for a technical definition) in the ARGs
categorised by parent node type, and demonstrates that each
of these contributes significantly to the overall topology.
In particular, we can see that approximately 40\% of the edges
in both ARGs involve inheriting directly from sampled genomes.
Both ARGs contain 29,422 sites (\ref{sec:data_preprocessing}),
and a large number of mutations.
The average number of mutations per-site is high, although much
of this may be explained by outlier sites with artefactually high
mutation counts (see Figure~\ref{fig:breakpoint-distribution}).
Despite this, however, the number of mutations per node
is small. In the Wide ARG, for example, we have a mean of 0.77
mutations per sample, demonstrating that new samples are generally
well represented by it. See Section~\ref{sec:mutation_spectrum}
for more analysis of the patterns of inferred mutations.
Recombination plays a relatively minor role, with $<0.3\%$
of nodes in the ARGs representing inferred recombination events.
We analyse these events in Sections~\ref{sec:jackson_recombs},
\ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},
and \ref{sec:pango_x_lineages}.

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing a local tree from the Wide ARG
(sampled to mid-2021) and an ``all-time'' global Nextstrain tree
(downloaded on January 21, 2023).
Phylogenies are pruned down to those samples present as tips in both datasets. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated using the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

A classic tree-based summary of SARS-CoV2 ancestry is provided
by the Nextstrain project \citep{Hadfield2018-ef}. The trees
available from Nextstrain are based on small subsamples of the
dataset, and early in the pandemic tend to be restricted to
a sample of strains which are not thought to be recombinants.
For validation purposes, we our compare our ARGs with
a downloaded Nextstrain tree, restricted to the time period
covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
the \texttt{sc2ts} ARGs to a backbone containing only samples present
the Nextstrain tree. This results in a small set of shared
samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
assigned to Pango recombinant lineages by Nextclade (see Section
\ref{sec:pango_x_lineages}).

Our \texttt{sc2ts} backbone phylogenies do contain small amounts of recombination:
there are 7 recombination nodes in the backbone of the Wide ARG and 8 in the backbone
of the Long ARG. However, the backbones remain extremely tree-like, as these
recombination events involve minor, local topological rearrangements. In particular,
recombination in the backbone phylogenies only occurs between close relatives
(e.g.\ sibling lineages), sometimes even within a single lineage such that the parent
lineages above a recombination event coalesce with each other before involvement with
any other part of the phylogeny. These local reticulations could either represent
real but previously unobserved recombination events between closely related strains,
or could be artifacts of the current \texttt{sc2ts} method and its parameterization.

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from Nextstrain
% We give the date below in the caption - we could probably remove if either from there
% or from here (I suggest from here)
(downloaded on January 21, 2023).
We illustrate the backbone phylogeny by visualizing
a single tree in the middle of the viral genome
since other regions of the
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the Nextstrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

\subsection{Early recombinants}
\label{sec:jackson_recombs}
% FIXME: this is rubbish, can someone do better?
RNA viruses are known to recombine at high rates 
\citep{Simon2011-rna}, and recombination has been widely 
documented to be commonplace in animal and human coronaviruses 
\citep{Su2016-epidemiology}. While recombination in SARS-CoV-2 
was shown early on to be frequent \emph{in-vitro} 
\citep{Gribble2021-coronavirus}, the relatively slow accumulation 
of genetic diversity early on in the pandemic hampered efforts to
detect recombinant strains. A number of early studies relying on 
analysing patterns of linkage disequilibrium and searching for 
mosaic genomes carrying characteristic mutations of different
lineages either failed to detect recombination or posited 
that this occurred at low rates \citep[e.g.~][]{Nie2020-phylogenetic,Tang2020-origin,VanInsberghe2021-eu,Varabyou2021-rw}.
The first clear evidence of recombinant lineages was presented by
\citet{Jackson2021-ik}, who performed a careful analysis of sequences
circulating in the UK in late 2020 to early 2021, and found
evidence of multiple independent recombination events and onward
transmission.
By searching for samples combining genomic segments from Alpha (B.1.1.7) and
from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
they found 16 recombinant sequences from 8 putative
origins (groups A to D and four singletons).
These findings are closely replicated in both the Wide and Long ARGs.

@@

Although recombination nodes are relatively infrequent, the wide and long ARGs still
contain many of them (4123 and 2078 respectively; Supplementary Tables 1 and 2).
Each of these nodes were
inferred as a result of matching a specific recombinant sample into the ARG.
However, ~63\% of recombination nodes are ancestral to only a single
recombinant sample; we exclude these "singleton recombinants" from subsequent
analyses as they may have arisen from sequencing of mixed samples (e.g., from
co-infection or contamination), and in any case, never result in onward
transmission.  This leaves 1522 and 763 recombination nodes in the wide and
long ARGs, respectively. In their analysis using RIPPLES,
\cite{Turakhia2022-it} detected 589 non-singleton recombinants.

@@

The date of common ancestry for the two parents of recombinant lineages is
concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
largely due to a small handful of MRCAs which are shared between a large number
of recombination events; the top four are indicated by dotted lines in the
figure. These shared MRCAs are near the root of large expansions of certain
lineages (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
and BA.2: second Omicron wave), as evidenced by the number of samples with those
Pango designations which trace back to each ancestral node (listed as
percentages on the plot). The majority of these frequently-shared MRCAs
are associated with extensive polytomies in the tree (at the extreme,
the BA.2-linked node has 2,872 immediate children). This is likely to
indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.

@@

\subsubsection{Wide ARG}
%% Quick overview of the wide ARG.
In the wide ARG, 44
samples are designated as XA by both Nextclade and GISAID, 237 are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
designated as XA, XB, or XC by any method have one or more recombination nodes
in their ancestry.

@@

The remaining remaining 630 Pango recombinant samples, representing 31
Nextclade-designated Pango lineages, trace back to 50 different recombination nodes.
We examine the inferred inheritance structure in the following subsections.

@@

A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
This shows a subgraph of the long ARG, in which paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset. The
subgraph may thus include inserted, non-sample nodes (i.e. UPGMA, recombination, mutation
collapse, or reversion push nodes, all plotted as smaller circles) on the path to the nearest
sample nodes (larger circles). Dotted lines show where this subgraph links to the rest of the
ARG. Above recombination nodes only ancestral nodes are shown, so that e.g. cousins of non-sample
recombinant parents are not shown. It is clear from the subgraph that all the XA samples trace
to a single originating
recombination node (plotted in black), with a breakpoint position around 22227bp.

@@

A more complex single-origin case is XAG, illustrated in
Figure~\ref{fig:pango-simple-origin-graph}B. Here, the XAG samples
all trace back to the same most recent common recombination node (black), but
we infer this recombination event to also be the originating event for
all the recombinant samples designated XAA,
and some, but not all, of those identified as XAB, XQ, and XU by Nextclade.

The classification of originating recombination events is dependent on
accurate designation of Pango lineages to samples. It is therefore important
to note that if the GISAID Pango designations are used, many of the samples
marked here as XAB are reclassified
as BA.2 (and XAG becomes fully monophyletic, see supplementary figure
\ref{fig:pango_XAG_gisaid_graph}).

In total we identify 16/31 single-origin Pango recombinant lineages in the long ARG.
Supplementary Table~\ref{tab:pango-single-origin} summarises these origination events
together with their inferred parents, showing good agreement between \texttt{sc2ts} and
the official Pango designated parents, with \texttt{sc2ts} providing more a precise
parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
designated lineages as descendants of the originating recombination event. In some cases,
these may be caused by samples which have been designated as the wrong recombinant lineage
by Nextclade.

\subsubsection{Multiple origins}
The ARG inference has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. In this section we examine the Pango X lineage
samples that are inferred to have multiple recombinant origins (i.e. trace
back to independent recombination nodes in the ARG).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL 8514045 (sampled in Turkey and dated 2021-12-30).
The different geographical locations, the time between the samples, and
the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent origins of XD.
Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs
with sample identifiers and coloured by GISAID Pango labels (however, no
samples are designated by GISAID as XD in the long ARG).

Another X lineage we infer to have multiple origins is XE. In this case, 
we identify at least 4 independent origination events of the Nextclade-designated
XE samples (Supplementary Table~\ref{tab:multiple_origins_table}).

[TODO: needs more detail]

\subsubsection{Nested recombinant origins}
As well as cases where single recombination events explain an X lineage,
such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
Pango X lineages involve more complex origins,
in which multiple non-independent (``nested'') recombination events are involved.
In these cases, the descendants of the original recombination
frequently include samples which have not been previously identified as recombinants.
An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are mostly not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
Note that the descendants of nested recombination events include only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Investigating examples of nested recombinant origins, and identifying which
(if any) of the nested recombination events may be artefactual, is an important
area of future research.
